Abstractions for Multi-Sorted Substitutions

### 2023-09-20 paper/paper.tex
Update.

    This paper describes an Agda framework for formalizing programming
    languages with extrinsic typing, intrinsic scoping, and de Bruijn
    Indices for variables.
    The framework supports object languages with arbitrary many
    variable sorts and dependencies, making it suitable for
    polymorphic languages and dependent types.
    Given an Agda definition of syntax and typing, the framework derives
    substitution operations and lemmas for untyped terms, and provides an
    abstraction to prove type preservation of these operations with
    just a single lemma.
    The key insights behind the framework are the use of multi-sorted syntax
    definitions, which enable parallel substitutions that replace
    all variables of all sorts simultaneously, and
    abstractions that unify the definitions, compositions, and typings
    of multi-sorted renamings and substitutions.
    Case studies have been conducted to prove subject reduction for
    System F with subtyping, dependently typed lambda calculus, and lambda
    calculus with pattern matching.
  \end{abstract}

@@

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservattion lemmas for renamings and substitutions.
  Section~\ref{sec:subjectreduction} introduces a small-step
  operational semantics for System F and proves subject reduction.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} discusses our case studies.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:future} discusses future work.
  Section~\ref{sec:conclusion} concludes.

  \subsection{Contributions}
  \label{sec:introduction:contributions}
  \begin{enumerate}
  \item
    a novel approach for formalizing intrinsically-scoped syntax with
    multiple variable sorts as a special kind of intrinsically-typed
    syntax, we call \emph{multi-sorted syntax};
  \item
    an adaptation of McBride's
    kit-abstraction\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
    for unifying renamings and substitutions to multi-sorted syntax;
  \item
    a novel abstraction for composition and it's metatheory, unifying
    the four compositions between renamings and substitutions;
  \item
    a novel abstraction for typing relations, unifying type
    preservation of renaming and substitution;
  \item
    a formalized specification of a large class of object languages
    for which substitution and lemmas can be derived generically.
  \item
    an implementation as an Agda framework featuring
    a reflection algorithm, representation independence for
    substitutions and typing contexts, heterogeneous equality between
    renamings and substitutions, and absence of postulated axioms.
  \item
    three case studies using our approach to prove subject reduction
    for System~F with subtyping, a dependently-typed lambda calculus, and
    pattern matching.
  \end{enumerate}

@@

  The operations need to satisfy the following properties:
  \begin{itemize}
  \item \ACode{\AField{`/`-is-`}} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \ACode{\AField{`\_}}.
  \item \ACode{\AField{`/id-injective}} and \ACode{\AField{id/`-injective}} state that
    \ACode{\AField{`/id}} and \ACode{\AField{id/`}} are injective. This property
    follows easily, as both the identity function and the variable
    constructor are injective.
  \item \ACode{\AField{wk-id/`}} characterizes the behaviour of the \ACode{\AField{wk}} function.
    Injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

@@

  As we defined the \ACode{\ACon{Kit}} structure before the
  \ACode{\ACon{Traversal}} structure, the fields of
  \ACode{\ACon{Kit}} could not use the \ACode{\ACon{\_⋯\_}} function in
  their types.

@@

  The core property of composition, which we call
  \ACode{\AField{⋯-fusion}}, states that applying two maps \ACode{ϕ₁}
  and \ACode{ϕ₂} in sequence to a term \ACode{t}, is the same as
  applying their composition \ACode{(ϕ₁ \AField{·ₖ} ϕ₂)} to \ACode{t},
  i.e.
  $$
  \ACode{(t \AField{⋯} ϕ₁) \AField{⋯} ϕ₂ \ACon{≡} t \AField{⋯} (ϕ₁ \AField{·ₖ} ϕ₂)}
  $$
  This property equips our framework with the power to
  reason about the application of multiple maps in terms of
  the application of a single map.
  As such it forms the basis for all lemmas involving multiple maps,
  e.g.\ that applying a weakening and then a singleton
  substitution cancel each other out.

@@

  Previous work on kits\cite{DBLP:journals/jar/BentonHKM12} addresses this
  issue by duplicating the definitions and using tactics to reduce
  boilerplate in proofs.
  In contrast, we define structures similar to \ACode{\ACon{Kit}} and
  \ACode{\ACon{Traversal}}, which allow us to abstract over all four
  compositions and use the same trick as before to eliminate the
  dependencies.
  This has the benefit, that we can prove lemmas about
  the composition of arbitrary maps, which enable us later to unify type
  preservation of all substitution and renaming operations into a single
  lemma.
  % TODO: clarify

@@

  \LibCode\KComposeKit

  Similar as for the \ACode{\ACon{Kit}}, we derive further definitions
  from the the axioms of a \ACode{\ACon{ComposeKit}}:
  \begin{itemize}
  \item
    We define map composition as motivated by the previous subsection:
    \LibCode\KComposition
  \item
    We show that the \ACode{\AField{\_\&/⋯\_}} operator behaves like
    \ACode{\AField{\_\&\_}} when applied to variables:
    \LibCode\KComposeKitAp
    % \LibCode\KComposeKitApProof

@@

  We then define the \ACode{\ACon{ComposeTraversal}} structure,
  which provides the \ACode{\ACon{⋯-fusion}} for the compositions
  described by arbitrary \ACode{\ACon{ComposeKit}}s:
  \LibCode\KComposeTraversal

  Similar as for the \ACode{\ACon{Traversal}}, we derive further
  definitions from the the axiom of a \ACode{\ACon{ComposeTraversal}}:
  \begin{itemize}
  \item 
    A map \ACode{ϕ} followed by a weakening is equivalent to a
    weakening followed by \ACode{ϕ} that has been lifted over the weakened
    variable:
    \LibCode\KCommLiftWeaken
    % \LibCode\KCommLiftWeakenProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the same
    statement for multiple map applications:
    \LibCode\KCommLiftWeakenTraverse
    % \LibCode\KCommLiftWeakenTraverseProof
  \item 
    A weakening followed by a singleton substitution cancel each other out:
    \LibCode\KWeakenCancelsSingle
    % \LibCode\KWeakenCancelsSingleProof
    By applying \ACode{\AField{⋯-fusion}} and \ACode{\AField{⋯-id}} to
    this lemma, we get the same statement for multiple map applications:
    \LibCode\KWeakenCancelsSingleTraverse
    % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton substitution followed by some map \ACode{ϕ} is the same
    as \ACode{ϕ} lifted over the variable of the singleton substitution
    followed by the singleton substitution with \ACode{ϕ} applied to
    its content:
    \LibCode\KDistLiftSingle
    % \LibCode\KDistLiftSingleProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the
    same statement for multiple map applications:
    \LibCode\KDistLiftSingleTraverse
    % \LibCode\KDistLiftSingleTraverseProof
  \item 
    Given the \ACode{\ADef{⋯-↑-wk}} lemma, we can define the four
    \ACode{\ACon{ComposeKit}} instances:
    \LibCode\KComposeKitInstances
    \LibCode\KComposeKitInstancesConcrete
  \end{itemize}

@@

  In the context of multi-sorted syntax, the notation of a type can be
  described as a mapping between sorts.
  For System F, the expression sort \ACode{\ACon{𝕖}} maps to the type sort
  \ACode{\ACon{𝕥}} and the sort \ACode{\ACon{𝕥}} maps to the kind sort
  \ACode{\ACon{𝕜}}.
  To teach our framework about types, we have to instantiate the following structure:
  \LibCode\KTypeSorts
  For System F the instantiation is
  \AppCode\FTypes
  For a nicer interface, we define a type analogously to a term:
  \LibCode\KTypes

@@

  \section{Reflection \& Generics}
  \label{sec:reflection}
  We used Agda's reflection mechanism\cite{TODO} to derive
  instantiations related to all structures for untyped substitition,
  i.e. \ACode{\ACon{Traversal}} and \ACode{\ACon{ComposeTraversal}}.

  To gain insight into the class of object languages supported by our
  reflection algorithm, we modeled a class of object languages using a
  generic programming technique and instantiated the structures
  generically. The class contains all object languages which:
  \begin{enumerate}
  \item Have a variable constructor of type TODO;
  \item Uses subterms only directly (e.g. not in lists);
  \item Subterms only extend the scope-context.
  \end{enumerate}

@@

  We believe restriction 3 can be lifted partially to allow for
  example the encoding of linear types that requires context
  splitting. This has already be done for intrinsical kits\cite{TODO}.

>>>>

    This paper describes an Agda framework for formalizing programming
    languages with extrinsic typing, intrinsic scoping, and de Bruijn
    Indices for variables.
    The framework supports object languages with arbitrary many
    variable sorts and dependencies, making it suitable for
    polymorphic languages and dependent types.
    Given an Agda definition of syntax and typing, the framework derives
    substitution operations and lemmas for untyped terms, and provides an
    abstraction to prove type preservation of these operations with
    just a single lemma.
    The key insights behind the framework are the use of multi-sorted syntax
    definitions, which enable parallel substitutions that replace
    all variables of all sorts simultaneously, and
    abstractions that unify the definitions, compositions, typings,
    and type preservation lemmas of multi-sorted renamings and substitutions.
    Case studies have been conducted to prove subject reduction for
    System F with subtyping, dependently typed lambda calculus, and lambda
    calculus with pattern matching.
  \end{abstract}

@@

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservation lemmas for renamings and substitutions.
  Section~\ref{sec:subjectreduction} introduces a small-step
  operational semantics for System F and proves subject reduction.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} discusses our case studies.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:future} discusses future work.
  Section~\ref{sec:conclusion} concludes.

  \subsection{Contributions}
  \label{sec:introduction:contributions}
  \begin{enumerate}
  \item
    a novel approach for formalizing intrinsically-scoped syntax with
    multiple variable sorts as a special kind of intrinsically-typed
    syntax, we call \emph{multi-sorted syntax};
  \item
    an adaptation of McBride's
    kit-abstraction\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
    for unifying renamings and substitutions to multi-sorted syntax;
  \item
    a novel abstraction for composition and it's metatheory, unifying
    the four compositions between renamings and substitutions;
  \item
    a novel abstraction for typing relations, unifying type
    preservation of renaming and substitution;
  \item
    a formalized specification of a large class of object languages
    for which untyped substitution and lemmas can be derived generically.
  \item
    an implementation as an Agda framework featuring
    a reflection algorithm, representation independence for
    substitutions and typing contexts, heterogeneous equality between
    renamings and substitutions, and absence of postulated axioms.
  \item
    three case studies in using our approach to prove subject reduction
    for System~F with subtyping, a dependently-typed lambda calculus, and
    pattern matching.
  \end{enumerate}

@@

  The operations need to satisfy the following properties:
  \begin{itemize}
  \item \ACode{\AField{`/`-is-`}} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \ACode{\AField{`\_}}.
  \item \ACode{\AField{`/id-injective}} and \ACode{\AField{id/`-injective}} state that
    \ACode{\AField{`/id}} and \ACode{\AField{id/`}} are injective. This property
    follows easily, as both the identity function and the variable
    constructor are injective.
  \item \ACode{\AField{wk-id/`}} characterizes the behaviour of the
    \ACode{\AField{wk}} function by how it acts on variables.
    Injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

@@

  As we defined the \ACode{\ACon{Kit}} structure before the
  \ACode{\ACon{Traversal}} structure, the fields of
  \ACode{\ACon{Kit}} could not use the \ACode{\AField{\_⋯\_}} function in
  their types.

@@

  The core property of composition, which we call
  \ACode{\AField{⋯-fusion}}, states that applying two maps \ACode{ϕ₁}
  and \ACode{ϕ₂} in sequence to a term \ACode{t}, is the same as
  applying their composition \ACode{(ϕ₁ \AField{·ₖ} ϕ₂)} to \ACode{t},
  i.e.
  $$
  \ACode{(t \AField{⋯} ϕ₁) \AField{⋯} ϕ₂ \ACon{≡} t \AField{⋯} (ϕ₁ \AField{·ₖ} ϕ₂)}
  $$
  This property equips our framework with the power to
  reason about the application of multiple maps by reasoning about
  the application of a single map.
  As such it forms the basis for all lemmas involving multiple maps,
  e.g.\ that applying a weakening and then a singleton
  substitution cancel each other out.

@@

  Previous work on kits\cite{DBLP:journals/jar/BentonHKM12} addresses this
  issue by duplicating the definitions and using tactics to reduce
  boilerplate in proofs.
  In contrast, we define structures similar to \ACode{\ACon{Kit}} and
  \ACode{\ACon{Traversal}}, which allow us to abstract over all four
  compositions and use the same trick as before to eliminate the
  dependencies.
  While composition can be dervied, this has the benefit, that
  we've gained a new level of generality, which enables us in the next
  sections to unify type preservation of all substitution and renaming
  operations into one lemma.

@@

  The \ACode{\ACon{ComposeKit}} abstraction provides the
  \ACode{\AField{\_\&/⋯\_}} operation, necessary for defining
  composition, and two lemmas about its behavior:
  \LibCode\KComposeKit
  The \ACode{\AField{\_\&/⋯\_}} operation takes a variable or term
  (according to \ACode{K₁}) and a renaming or substitution
  (according to \ACode{K₂}) and performs the map traversal returning
  a variable or term (according to \ACode{K₁⊔K₂}).
  We later define four instances for \ACode{\ACon{ComposeKit}}, where
  \ACode{K₁} and \ACode{K₂} can each be either the variable or term kit.
  The third parameter \ACode{K₁⊔K₂} can be seen as a functional
  dependency and is completely determined by the choice of \ACode{K₁}
  and \ACode{K₂}.

  The \ACode{\AField{\&/⋯-⋯}} lemma describes the behavior of
  \ACode{\AField{\_\&/⋯\_}} in terms of \ACode{\AField{\_⋯\_}}, allowing
  subsequent lemmas to make use of the lemmas we previously proved
  about \ACode{\AField{\_⋯\_}}.

  The \ACode{\AField{\&/⋯-wk-↑}} lemma states that applying a map and then
  weakening is the same as weakening first and then lifting the map over
  the variable introduced by the weakening.

  Similar as for the \ACode{\ACon{Kit}}, we derive further definitions
  from the axioms of a \ACode{\ACon{ComposeKit}}:
  \begin{itemize}
  \item
    We define map composition as motivated by the previous subsection:
    \LibCode\KComposition
  \item
    We show that the \ACode{\AField{\_\&/⋯\_}} operator behaves like
    \ACode{\AField{\_\&\_}} when applied to variables:
    \LibCode\KComposeKitAp
    % \LibCode\KComposeKitApProof

@@

  We then define the \ACode{\ACon{ComposeTraversal}} structure,
  which provides the \ACode{\AField{⋯-fusion}} lemma for the compositions
  described by arbitrary \ACode{\ACon{ComposeKit}}s:
  \LibCode\KComposeTraversal

  Similar as for the \ACode{\ACon{Traversal}}, we derive further
  definitions from the axiom of a \ACode{\ACon{ComposeTraversal}}:
  \begin{itemize}
  \item 
    A map \ACode{ϕ} followed by a weakening is equivalent to a
    weakening followed by \ACode{ϕ} that has been lifted over the weakened
    variable:
    \LibCode\KCommLiftWeaken
    % \LibCode\KCommLiftWeakenProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the same
    statement for multiple map applications:
    \LibCode\KCommLiftWeakenTraverse
    % \LibCode\KCommLiftWeakenTraverseProof
  \item 
    A weakening followed by a singleton substitution cancel each other out:
    \LibCode\KWeakenCancelsSingle
    % \LibCode\KWeakenCancelsSingleProof
    By applying \ACode{\AField{⋯-fusion}} and \ACode{\AField{⋯-id}} to
    this lemma, we get the same statement for multiple map applications:
    \LibCode\KWeakenCancelsSingleTraverse
    % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton substitution followed by some map \ACode{ϕ} is the same
    as \ACode{ϕ} lifted over the variable of the singleton substitution
    followed by the singleton substitution with \ACode{ϕ} applied to
    its content:
    \LibCode\KDistLiftSingle
    % \LibCode\KDistLiftSingleProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the
    same statement for multiple map applications:
    \LibCode\KDistLiftSingleTraverse
    % \LibCode\KDistLiftSingleTraverseProof
  \item 
    Given the \ACode{\ADef{⋯-↑-wk}} lemma, we can define the four
    \ACode{\ACon{ComposeKit}} instances:
    \LibCode\KComposeKitInstances
    \LibCode\KComposeKitInstancesConcrete
  \end{itemize}

@@

  In the context of multi-sorted syntax, the notion of a type can be
  described as a mapping between sorts.
  For System F, the expression sort \ACode{\ACon{𝕖}} maps to the type sort
  \ACode{\ACon{𝕥}} and the sort \ACode{\ACon{𝕥}} maps to the kind sort
  \ACode{\ACon{𝕜}}.
  To teach our framework about types, we have to instantiate the following structure:
  \LibCode\KTypeSorts
  For System F the instantiation is
  \AppCode\FTypes
  For a nicer interface, we define a type analogously to a term:
  \LibCode\KTypes

@@

  \section{Reflection \& Generics}
  \label{sec:reflection}
  We used Agda's reflection mechanism\cite{TODO} to derive
  instantiations related to all structures for untyped substitition,
  i.e. \ACode{\ACon{Syntax}}, \ACode{\ACon{Traversal}} and
  \ACode{\ACon{ComposeTraversal}}.

  To gain insight into the class of object languages supported by our
  reflection algorithm, we have formalized a class of object languages
  using a generic programming technique and instantiated the structures
  generically. Informally, this class contains all object languages which
  \begin{enumerate}
  \item have a variable constructor of type TODO;
  \item use subterms only directly (e.g. not in lists);
  \item only extend the scope-context of subterms.
  \end{enumerate}

@@

  We believe restriction 3 can be lifted partially to allow for
  example the encoding of linear types that requires context
  splitting.
